---
title: "Introduction to SIRA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SIRA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

SIRA (Scalable Image-on-Scalar Regression Algorithm) fits image-on-scalar
regression models of the form:

$$Y_i(s_v) = \sum_{j=1}^{p_1} \beta_j(s_v) X_{ij} + \sum_{k=1}^{p_2} \gamma_k(s_v) Z_{ik} + \eta_i(s_v) + \epsilon_i(s_v)$$

where:

- $Y_i(s_v)$ is subject $i$'s image value at voxel $v$
- $X_{ij}$ are covariates of interest with spatially-varying coefficient images $\beta_j$
- $Z_{ik}$ are confounders (typically including an intercept) with coefficient images $\gamma_k$
- $\eta_i(s_v) = \theta_i \Psi^*(s_v)$ is a spatial individual effect
- $\epsilon_i(s_v)$ is noise

SIRA enforces two properties on each $\beta_j$: **sparsity** (most voxels have
$\beta_j = 0$) and **spatial homogeneity** (nonzero voxels form spatially
contiguous regions with a constant value). It achieves this via a
coordinate-descent algorithm with five region operations: revalue, expand,
shrink, merge, and split.

## A simple example

We begin with a single covariate and a small 10×10×10 image to verify basic
functionality.

```{r simple-setup}
library(SIRA)

set.seed(1)
d1 <- d2 <- d3 <- 10L
V  <- d1 * d2 * d3
n  <- 50L

# True signal: a 3x3x3 cube of voxels near the origin with beta = 3
beta_true        <- numeric(V)
beta_true[1:27]  <- 3

X <- matrix(rnorm(n), ncol = 1)
Z <- cbind(1, rnorm(n))
Y <- X %*% t(beta_true) +
     Z %*% matrix(rnorm(2 * V), nrow = 2) +
     matrix(rnorm(n * V, sd = 3), nrow = n)
```

```{r simple-fit}
fit <- sira(Y = Y, X = X, Z = Z,
            d1 = d1, d2 = d2, d3 = d3,
            lambda = 0.05, mu = 0.01)
print(fit)
```

We can check how well SIRA recovered the true signal region:

```{r simple-check}
recovered   <- which(fit$betahat[1, ] != 0)
true_voxels <- which(beta_true != 0)

cat("True positives :", length(intersect(recovered, true_voxels)),
    "of", length(true_voxels), "\n")
cat("False positives:", length(setdiff(recovered, true_voxels)), "\n")
```

With a small sample (n=50) and high noise (sd=3), SIRA recovers all 27 true
signal voxels with no false positives.

## A more challenging example

Here we fit a model with two covariates, each having two spatially separated
regions with opposite signs. This tests that SIRA correctly handles multiple
covariates, positive and negative regions, and that the two coefficient images
do not interfere with each other.

```{r hard-setup}
set.seed(42)
n <- 200L

# Helper: linear voxel indices for a 3D box in column-major order
box <- function(i1r, i2r, i3r) {
  as.integer(unlist(lapply(i3r, function(k)
    unlist(lapply(i2r, function(j)
      i1r + d1 * (j - 1L) + d1 * d2 * (k - 1L))))))
}

# Beta 1: positive cube near (1,1,1), negative cube near (8,8,8)
# Beta 2: positive cube near (1,8,1), negative cube near (8,1,8)
# All four regions are fully separated (gap >= 4 voxels on every axis).
regions <- list(
  b1_pos = list(vox = box(1:3, 1:3,   1:3),   val =  2.5),
  b1_neg = list(vox = box(8:10, 8:10, 8:10),  val = -2.5),
  b2_pos = list(vox = box(1:3, 8:10,  1:3),   val =  2.0),
  b2_neg = list(vox = box(8:10, 1:3,  8:10),  val = -2.0)
)

beta_true2 <- matrix(0, nrow = 2, ncol = V)
beta_true2[1, regions$b1_pos$vox] <-  2.5
beta_true2[1, regions$b1_neg$vox] <- -2.5
beta_true2[2, regions$b2_pos$vox] <-  2.0
beta_true2[2, regions$b2_neg$vox] <- -2.0

X2 <- matrix(rnorm(n * 2), ncol = 2)
Z2 <- cbind(1, rnorm(n))
Y2 <- X2 %*% beta_true2 +
      Z2 %*% matrix(rnorm(2 * V), nrow = 2) +
      matrix(rnorm(n * V, sd = 2), nrow = n)
```

```{r hard-fit}
fit2 <- sira(Y = Y2, X = X2, Z = Z2,
             d1 = d1, d2 = d2, d3 = d3,
             lambda = 0.05, mu = 0.01)
print(fit2)
```

```{r hard-check}
check_covariate <- function(j, true_pos, true_neg) {
  bhat     <- fit2$betahat[j, ]
  rec      <- which(bhat != 0)
  pos_rec  <- rec[bhat[rec] > 0]
  neg_rec  <- rec[bhat[rec] < 0]
  all_true <- c(true_pos, true_neg)

  cat(sprintf("Covariate %d:\n", j))
  cat(sprintf("  Regions found   : %d  (expected 2)\n",
              length(fit2$region_list[[j]])))
  cat(sprintf("  Positive region : %d voxels, mean beta = %.3f  (true beta = %.1f)\n",
              length(pos_rec),
              if (length(pos_rec) > 0) mean(bhat[pos_rec]) else NA,
              beta_true2[j, true_pos[1]]))
  cat(sprintf("  Negative region : %d voxels, mean beta = %.3f  (true beta = %.1f)\n",
              length(neg_rec),
              if (length(neg_rec) > 0) mean(bhat[neg_rec]) else NA,
              beta_true2[j, true_neg[1]]))
  cat(sprintf("  True positives  : %d / %d (%.0f%%)\n",
              length(intersect(rec, all_true)), length(all_true),
              100 * length(intersect(rec, all_true)) / length(all_true)))
  cat(sprintf("  False positives : %d\n\n",
              length(setdiff(rec, all_true))))
}

check_covariate(1, regions$b1_pos$vox, regions$b1_neg$vox)
check_covariate(2, regions$b2_pos$vox, regions$b2_neg$vox)
```

Both coefficient images are recovered with the correct region structure (2
regions each), correct signs, and no false positives. The estimated beta values
are attenuated relative to the truth due to the L1 penalty (`mu`) and partial
absorption of signal by the spatial random effects term — this is expected
behavior and can be reduced by lowering `mu`.

## Reading NIfTI batch files

`sira_batched()` accepts any file format via the `Y_reader` argument. For
NIfTI files (`.nii` or `.nii.gz`), each batch file is typically stored as a
4D array (d1 × d2 × d3 × n_subjects). The reader needs to reshape this into
the n_b × V matrix that `sira_batched()` expects:

```{r nifti-example, eval = FALSE}
library(RNifti)

# Custom reader: reads a 4D NIfTI and returns an n_b x V matrix
nifti_reader <- function(path) {
  img <- readNifti(path)                      # d1 x d2 x d3 x n_b array
  n_b <- dim(img)[4]
  V   <- prod(dim(img)[1:3])
  matrix(aperm(img, c(4, 1, 2, 3)), nrow = n_b, ncol = V)
}

# Each .nii.gz file contains one batch of subjects
fit <- sira_batched(
  Y_files  = list.files("path/to/batches", pattern = "\\.nii\\.gz$",
                        full.names = TRUE),
  Y_reader = nifti_reader,
  X = X, Z = Z,
  d1 = d1, d2 = d2, d3 = d3,
  lambda = 0.05, mu = 0.01
)
```

For brain-masked data where each NIfTI contains only the voxels within a mask,
pair this with the `coords` argument instead of `d1/d2/d3`:

```{r nifti-masked-example, eval = FALSE}
# coords is a V x 3 matrix of (x, y, z) coordinates for the masked voxels
fit_masked <- sira_batched(
  Y_files  = list.files("path/to/batches", pattern = "\\.nii\\.gz$",
                        full.names = TRUE),
  Y_reader = nifti_reader,
  X = X, Z = Z,
  coords = mni_coords,
  lambda = 0.05, mu = 0.01
)
```

Install `RNifti` with `install.packages("RNifti")`. Other NIfTI readers such
as `oro.nifti::readNIfTI()` or `neurobase::readnii()` work equally well —
just adjust the reader function to return an n_b × V numeric matrix.

## Extracting results

The fitted coefficient images can be extracted and reshaped into 3D arrays for
visualization:

```{r extract}
# Extract the p1 x V betahat matrix
beta_mat <- coef(fit2, type = "beta")   # 2 x 1000

# Reshape row 1 into a 10 x 10 x 10 array
beta1_array <- as.array(fit2, j = 1)
dim(beta1_array)

# A simple slice through the first axis
beta1_array[1, , ]
```

## Irregular and brain-masked data

When voxels do not sit on a rectangular grid — for example, when `Y` contains
only grey-matter masked voxels from a brain atlas — pass a V x 3 matrix of
spatial coordinates instead of `d1`, `d2`, `d3`:

```{r coords-example, eval = FALSE}
# coords is a V x 3 matrix of (x, y, z) MNI coordinates (in mm or voxels)
fit_masked <- sira(Y = Y_masked, X = X, Z = Z,
                   coords = mni_coords,
                   lambda = 0.05, mu = 0.01)
```

SIRA determines adjacency automatically: two voxels are 6-connected neighbours
if they differ by exactly one voxel spacing along one axis and are coincident
on the other two. The spacing is detected from the minimum gap between unique
coordinate values on each axis, so both millimetre and voxel-index coordinates
work without any manual scaling.

All downstream code (`print`, `coef`, `summary`) works identically. The only
difference is that `as.array()` is not available for masked fits (since there
is no rectangular grid to reshape into), so use `coef()` to extract the
length-V coefficient vector and map it back to your atlas using your own
masking index.

## Tuning parameters

SIRA has two tuning parameters:

- **`lambda`** controls spatial smoothness. Larger values encourage larger,
  blockier regions by penalizing boundaries more heavily.
- **`mu`** controls sparsity. Larger values produce fewer nonzero voxels.

Both can be selected by cross-validation or by inspecting the convergence
behavior across a grid of values.
